A numerical study of the solution of X-Mode equations around the hybrid resonance

### 2015-01-29 report/abstract.tex
Bruno's abstract/resume/intro/FD/timedomain

Hybrid resonnance is a physical phenomenon that appears for example in the heating of plasma, and as such is of interest in the developping of new energy source, like for instance in the ITER project. In this paper we focus on the  non-regularity of
the solutions of Maxwell equations in plasmas under strong background magnetic field. Our purpose here is two-fold. On one hand we investigate the finite element approximation of the one dimensionnal problem written in the frequency domain, and on the other hand we investigate the finite difference approximation of the one dimensionnal time dependant problem. We will also compare the results of theses to differents methods.

>>>>

Hybrid resonance is a physical phenomenon
that appears for example in the heating of plasma,
and as such is of scientific interest in the development  of  the ITER project.
In this paper we focus some solutions with low regularity of Maxwell equations in
plasmas under strong background magnetic field.
Our purpose is two-fold.
On one hand we investigate the finite element approximation of the one dimensional problem 
written in the frequency domain, and on the other hand we investigate two different 
finite difference approximations of the one dimensional time dependent problem.
We will also compare the results of these different methods.

### 2015-01-29 report/intro.tex
Bruno's abstract/resume/intro/FD/timedomain

We consider the model of cold plasma of \cite{Stix} which is described by the 2D time-dependent Maxwell system  
\bealn
-\varepsilon_0 \partial_t \E + \curl\, \Hbf = \J\\
\mu_0 \partial_t \Hbf + \curl\, \E = 0
\eealn
coupled with a linear electronic current  $\J = eN_e \ubf_e$, 
\ben
m_e \partial_t \ubf_e =e (\E +\ubf_e \wedge B_0) -m_e \nu \ubf_e, \label{eq:electronmove}
\een
where $\nu$ is  the friction  between particles.
The unknowns are the electromagnetic fields $(\E,\Hbf)$ with the usual notation $\Hbf = \B/\mu_0$, 
 %the electronic current $\J$, 
and the velocity of electrons $\ubf_e$. Here $B_0$ is the background magnetic field, typically assumed to be uniform in time and space.  
We denote by $e<0$ the value of the charge of electrons, $m_e$ the electron mass and $N_e$ the electron density that in general 
depends on spatial variables and is uniform in time. 
Without loss of generality, we set $\mathbf{B}_0=\left(0,\; 0,\; B_0\right)$, which allows to obtain the following system of equations 
\begin{align}
\label{eq:main_model}
\left\{\begin{array}{l}
\epsilon_0\partial_t E_{x}-\partial_y H_z=-eN_e u_x,\\
\epsilon_0\partial_t E_{y}+\partial_x H_z=-eN_e u_y,\\
\mu_0\partial_t H_z+\partial_x E_y-\partial_y E_x=0,\\
m_e\partial_t u_x=eE_x+eu_yB_0-\nu m_e u_x,\\
m_e\partial_t u_y=eE_y-eu_xB_0-\nu m_e u_y
\end{array}
\right.\end{align}
posed in $[-L,\; \infty)\times \mathbb{R}$, for some $L>0$. This domain represents the physical case of a wave sent from a wall facing a plasma. 
The energy of this system for $\nu=0$ in a domain $\Omega\in\mathbb R^2$ can be expressed as \cite{stable_yee_plasma_current}
\mrev{\ben
{\mathcal E}(t)= \int_\Omega \left(
\frac{\epsilon_0 |\E(t,\x)|^2}{2}+\frac{ |\B(t,\x)|^2}{2\mu_0}+\frac{m_e N_e|\mathbf{u} (t,\x)|^2}{2} 
\right)\mathrm{d}\x.
\een}
\mrev{An energy-preserving scheme for discretizing (\ref{eq:main_model}) was recently developed in \cite{Despres_2014}. 
It can be shown that for non-zero collision frequency $\nu$ the loss of energy is proportional to $\nu$, however, as $\nu\rightarrow0+$, it 
does not vanish. This behaviour had been recently investigated in \cite{Despres_2014}}.

We complement Problem (\ref{eq:main_model}) with the following boundary conditions.
At the left boundary of the domain, which represents the wall, we choose Robin boundary conditions 
\ben
-curl \E -\imath \lambda\E \wedge \n = g_{inc} = -curl \E_{inc} -\imath\lambda\E_{inc} \wedge \n,
\een
where $\E_{inc} = \exp \left(\imath \lambda x\right)\begin{pmatrix} E_1\\ E_2 \end{pmatrix}$. We truncate the domain 
$(-L,\; \infty)$ to $(-L,\; H)$ and set on the right boundary $\operatorname{curl} \E = 0$. 
\begin{remark}
	When modelling antennas in plasma physics \mrev{(e.g. in such applications as ITER)}, 
	a good choice of boundary condition could be $\lambda = \sqrt{\alpha(-L)}$
\end{remark}

The goal of this article is two-fold. 
First, we investigate the finite element approximation of the 1D 
problem \eqref{eq:main_model} written in the frequency domain ($\partial_t\rightarrow -i\omega$), with a linearized dielectric tensor with respect to $\nu$.
In particular, we again investigate this problem in one dimension, performing the Fourier transform in $y$.
We prove the well-posedness of this problem for $\nu>0$ in Section~\ref{sec:wellposedness} and 
demonstrate that the use of higher-order (P1) finite elements allows to approximate the singularity 
of the solution fairly well (Section \ref{sec:freq_dep}). 
Second, we consider the case $\nu\rightarrow 0$, and study the limiting amplitude solution 
$\lim\limits_{t\rightarrow +\infty}\lim\limits_{\nu\rightarrow 0}\mathbf{E}(t)$ obtained with the help of 
the FDTD discretization of \eqref{eq:main_model}, suggested in \cite{stable_yee_plasma_current}. 
We compare this result with 
$\hat{\mathbf{E}}\mathrm{e}^{i\omega t}$, computed in the frequency domain, for $\nu\rightarrow 0$, which is equivalent to considering
$\lim\limits_{\nu\rightarrow 0}\lim\limits_{t\rightarrow+\infty}\mathbf{E}(t)$, i.e. we check the validity of the scheme in Figure \ref{fig:limits}. 
\begin{figure}
\begin{tikzpicture}
   \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em] {
     \mathbf{E}^{\nu}(t) & \mathbf{E}^{0+}(t)\\
     \hat{\mathbf{E}}^{\nu}(t)\mathrm{e}^{i\omega t} & 
     \lim\limits_{\nu\rightarrow 0}\lim\limits_{t\rightarrow \infty}\mathbf{E}^{\nu}(t)=\lim\limits_{t\rightarrow \infty}\lim\limits_{\nu\rightarrow 0}\mathbf{E}^{\nu}(t)\\
     };
  \path[-stealth,font=\scriptsize]    (m-1-1) edge node [left] {$t\rightarrow\infty$} (m-2-1);
  \path[-stealth, font=\scriptsize]   (m-1-1) edge node [above] {$\nu\rightarrow 0+$} (m-1-2);
  \path[-stealth, font=\scriptsize]   (m-2-1) edge node [above] {$\nu\rightarrow 0$} (m-2-2);
  \path[-stealth, font=\scriptsize]   (m-1-2) edge node [right] {$t\rightarrow \infty$} (m-2-2);

  % \path[-stealth] (m-2-1.east) edge node [above] {$\nu\rightarrow 0$} (m-2-2);
    %(m-1-2) edge node [right] {$\nu\rightarrow 0$} (m-2-2);
\end{tikzpicture}
 \caption{\label{fig:limits}Schematic representation of the equivalence of the limited absorption and limiting amplitude principles.}

\end{figure}

>>>>

We consider the model of cold plasma of \cite{Stix} which is described by the 2D time-dependent Maxwell system  
\bealn
-\varepsilon_0 \partial_t \E + \curl\, \Hbf = \J\\
\mu_0 \partial_t \Hbf + \curl\, \E = 0
\eealn
coupled with a linear electronic current  $\J = eN_e \ubf_e$, 
\ben
m_e \partial_t \ubf_e =e (\E +\ubf_e \wedge B_0) -m_e \nu \ubf_e, \label{eq:electronmove}
\een
where $\nu$ is  the friction  between particles.
The unknowns are the electromagnetic fields $(\E,\Hbf)$ with the usual notation $\Hbf = \B/\mu_0$, 
 %the electronic current $\J$, 
and the velocity of electrons $\ubf_e$. Here $B_0$ is the background magnetic field, typically assumed to be uniform in time and space.  
We denote by $e<0$ the value of the charge of electrons, $m_e$ the electron mass and $N_e$ the electron density that in general 
depends on spatial variables and is uniform in time. 
Without loss of generality, we set $\mathbf{B}_0=\left(0,\; 0,\; B_0\right)$, which allows to obtain the following system of 1D equations 
\begin{align}
\label{eq:main_model}
\left\{\begin{array}{l}
\epsilon_0\partial_t E_{x}&=-eN_e u_x,\\
\epsilon_0\partial_t E_{y}+\partial_x H_z&=-eN_e u_y,\\
\mu_0\partial_t H_z+\partial_x E_y &=0,\\
m_e\partial_t u_x&=eE_x+eu_yB_0-\nu m_e u_x,\\
m_e\partial_t u_y&=eE_y-eu_xB_0-\nu m_e u_y
\end{array}
\right.\end{align}
posed in $[-L,\; \infty)\times \mathbb{R}$, for some $L>0$. This domain represents the physical case of a wave sent from a wall facing a plasma. 
The energy of this system for $\nu=0$ in a domain $\Omega\in\mathbb R^2$ can be expressed as \cite{stable_yee_plasma_current}
\mrev{\ben
{\mathcal E}(t)= \int_\Omega \left(
\frac{\epsilon_0 |\E(t,\x)|^2}{2}+\frac{ |\B(t,\x)|^2}{2\mu_0}+\frac{m_e|\J (t,\x)|^2}{2|e|N_e(\x)} 
\right)\mathrm{d}\x.
\een}
It can be shown    \cite{Despres_2014} 
that the time harmonic electric field component $E_x$ in this case is no longer square 
integrable, with 
explicit estimates on the behavior of the solutions of \eqref{eq:main_model} in 1D were provided. 
This apparent paradox  is of course the source of important numerical difficulties which are the subject of the present study.

We complement Problem (\ref{eq:main_model}) with the boundary conditions.
At the left boundary of the domain, which represents the wall of the Tokamak, we choose Robin boundary conditions 
\be
-curl \E -\imath \omega\E \wedge \n = g_{inc} = -curl \E_{inc} -\imath\omega\E_{inc} \wedge \n,
\ee
where $\E_{inc} = \exp \left(\imath \omega x\right)\begin{pmatrix} E_1\\ E_2 \end{pmatrix}$ and $\omega$ is 
the frequency of the antenna.
%some given 
%frequency. 
We truncate the domain 
\ben
-curl \E -\imath \lambda\E \wedge \n = g_{inc} = -curl \E_{inc} -\imath\lambda\E_{inc} \wedge \n,
\een
where $\E_{inc} = \exp \left(\imath \lambda x\right)\begin{pmatrix} E_1\\ E_2 \end{pmatrix}$. We truncate the domain 
$(-L,\; \infty)$ to $(-L,\; H)$ and set on the right boundary $\operatorname{curl} \E = 0$. 
%\begin{remark}
%	When modelling antennas in plasma physics \mrev{(e.g. in such applications as ITER)}, 
%	a good choice of boundary condition could be $\lambda = \sqrt{\alpha(-L)}$
%\end{remark}

The goal of this article is three-fold. 
First, we investigate the finite element approximation of the 1D 
problem \eqref{eq:main_model} written in the frequency domain ($\partial_t\rightarrow -i\omega$), with a linearized dielectric tensor with respect to $\nu$.
In particular, we again investigate this problem in one dimension, performing the Fourier transform in $y$.
We prove the well-posedness of this problem for $\nu>0$ in Section~\ref{sec:wellposedness} and 
demonstrate that the use of  (P1) finite elements allows to approximate the singularity 
of the solution fairly well (Section \ref{sec:freq_dep}). 
Second, we briefly develop an original scheme based on widely appreciated semi-lagrangian schemes
for the discretization of time domain Maxwell's equations with a linear current.
Thirdly we consider the case $\nu\rightarrow 0$, and study the limiting amplitude solution 
$\lim\limits_{t\rightarrow +\infty}\lim\limits_{\nu\rightarrow 0}\mathbf{E}(t)$ obtained with the help of 
the FDTD discretization of \eqref{eq:main_model}, suggested in \cite{stable_yee_plasma_current}. 
We compare this result with 
$\hat{\mathbf{E}}\mathrm{e}^{i\omega t}$, computed in the frequency domain, for $\nu\rightarrow 0$, which is equivalent to considering
$\lim\limits_{\nu\rightarrow 0}\lim\limits_{t\rightarrow+\infty}\mathbf{E}(t)$.
Such a comparison is a way to study the formal commutation relation
$$
\lim\limits_{\nu\rightarrow 0}\lim\limits_{t\rightarrow+\infty}= \lim\limits_{t\rightarrow+\infty}\lim\limits_{\nu\rightarrow 0}.
$$
Even it is true for standard linear wave problems, it is not clear that it still hold in our case due the singularity
of the hybrid resonance.
%\begin{figure}
%\begin{tikzpicture}
%   \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em] {
%     \mathbf{E}^{\nu}(t) & \mathbf{E}^{0+}(t)\\
%     \hat{\mathbf{E}}^{\nu}(t)\mathrm{e}^{i\omega t} & 
%     \lim\limits_{\nu\rightarrow 0}\lim\limits_{t\rightarrow \infty}\mathbf{E}^{\nu}(t)=\lim\limits_{t\rightarrow \infty}\lim\limits_{\nu\rightarrow 0}\mathbf{E}^{\nu}(t)\\
%     };
%  \path[-stealth,font=\scriptsize]    (m-1-1) edge node [left] {$t\rightarrow\infty$} (m-2-1);
%  \path[-stealth, font=\scriptsize]   (m-1-1) edge node [above] {$\nu\rightarrow 0+$} (m-1-2);
%  \path[-stealth, font=\scriptsize]   (m-2-1) edge node [above] {$\nu\rightarrow 0$} (m-2-2);
%  \path[-stealth, font=\scriptsize]   (m-1-2) edge node [right] {$t\rightarrow \infty$} (m-2-2);
%
%
%  % \path[-stealth] (m-2-1.east) edge node [above] {$\nu\rightarrow 0$} (m-2-2);
%    %(m-1-2) edge node [right] {$\nu\rightarrow 0$} (m-2-2);
%\end{tikzpicture}
% \caption{\label{fig:limits}Schematic representation of the equivalence of the limited absorption and limiting amplitude principles.}
%
%\end{figure}

### 2015-01-29 report/resume.tex
Bruno's abstract/resume/intro/FD/timedomain

La r\'esonnance hybride est un ph\'enom\`ene physique qui apparait par exemple lorsque l'on chauffe un plasma, et est ainsi d'int\'er\^et dans le d\'eveloppement d'une nouvelle source d'\'energie, comme dans le cadre du projet ITER. Dans ce papier, nous nous concentrons sur la non r\'egularit\'e des solutions des \'equations de Maxwell pour les plasmas sous l'influence de champ magn\'etiques forts. Notre but est ici double. D'un c\^ot\'e nous regardons l'approximation \`a l'aide d'\'el\'ements finis du probl\`eme en une dimension \'ecrit dans le domaine en fr\'equence, et de l'autre nous regardons l'approximation \`a l'aide de diff\'erences finies du probl\`eme en une dimension mais d\'ependant en temps. Nous comparons \'egalement les r\'esultats de ces deux diff\'erentes m\'ethodes.

>>>>

La r\'esonnance hybride est un ph\'enom\`ene physique qui apparait par exemple lorsque l'on chauffe un plasma, et 
ainsi est d'int\'er\^et scientifique dans le cadre du d\'eveloppement  du projet ITER. Dans ce papier, nous nous concentrons sur certaines  solutions faiblement r\'eguli\`eres des \'equations de Maxwell pour les plasmas sous l'influence de champ magn\'etiques forts. Notre but est ici double. D'un c\^ot\'e nous \'evaluons l'approximation 
num\'erique \`a l'aide d'\'el\'ements finis en une dimension en formulation 
fr\'equentielle, et de l'autre nous \'etudions l'approximation num\'erique \`a l'aide de deux m\'ethodes de  diff\'erences finies pour la formulation
temporelle monodimensionelle. Nous comparons les r\'esultats de ces  diff\'erentes m\'ethodes.
